# Basic `R` Usage

```{r, message=FALSE, warning=FALSE}
library("swirl")
library("rio")
library("ggplot2")
```

The `R` programming language and software environment for statistical computing is an implementation of the proprietary `S` programming language by Ross Ithaka and Robert Gentleman in 1992. It quickly gained in popularity (see, e.g, this Nature article from 2014 [Programming tools: Adventures with R](http://www.nature.com/news/programming-tools-adventures-with-r-1.16609)) and now has more than 8,000 user contributed packages (see as well the blog piece [On the growth of CRAN packages](https://www.r-bloggers.com/on-the-growth-of-cran-packages/)).

## Where to get it

First, you need to install `R` itself. You can do this by downloading this from [CRAN](http://cran.xl-mirror.nl/) (we choose here the server from the Netherlands). Choose your appropriate operating system, choose the `base` system, download `R` and install it. That's it!

The base distribution of `R` comes with a built-in editor, where you can write your script (more about scripts in subsection \@ref(subsec:scripts)). This editor is however *very* basic. Therefore, it is very much advised that you download and install the Open Source editor [`RStudio`](https://www.rstudio.com/products/rstudio/download/) as well. Again, choose your operating system and just install the latest version. The very short video (1.5 minutes) on RStudio's [website](https://www.rstudio.com/products/rstudio/) gives an overview of the *basic* features (it can do quite some more stuff). 

## For absolute beginners

At the moment there are *many* tutorials, blogs, youtube clips, and background materials about using `R` on the internet. I therefore do not intend to write a complete handbook, but focus instead on what I need for my courses. Moreover, I do not intend to teach the very basic stuff. There is very good tutorial package out there called Swirl (see as well the website: [http://swirlstats.com/students.html](http://swirlstats.com/students.html)). I very much recommend using this package for absolute beginners. The way to do this is rather simple. First install the package by typing:

```{r, eval=FALSE}
install.packages("swirl")
```

then start Swirl by first loading the package:

```{r, eval=FALSE}
library("swirl")
```

And then call the function by typing: 

```{r, eval=FALSE}
swirl()
```

In the first menu choose `R Programming`. Now, there are 15 lessons. I find the first four the most useful (Basic Building Block, Workspace and Files, Sequences of Numbers, Vectors), but others are very useful as well to go through. The command `main()` by the way brings you always back to the main menu, and do not go for the credits on coursera (that is now a paid online course).

## How to use it

`R` is truly a programming language in the sense that there is no graphical user interface (GUI) involved. You need to type your own commands. And for beginners this sort of sucks. It seems slower, you have no idea which commands to type in, and you frequently make many mistakes. However, when you start to use it more, speed of getting things done goes up (sometimes exponentially), you have a better grasp on the basic commands, and the number of mistakes go down. In contrast with programs such as `Excel` or `SPSS`, there are two big differences: (*i*) you use scripts and (*ii*) you make frequently use of packages which are essentially written by other `R` users. 

### Scripts {#subsec:scripts}

The use of scripts or program files is somewhat alien to most. Although `Stata` also makes use of so-called `do` files. You start a new script by clicking on File > New Script (`R` editor) or File > New File > New Script (`RStudio`). You now have a new *empty* file (which you have to save from time to time). If you fill in this file with commands, you are actually programming. The **huge** benefit of this procedure is that you record what you have done and that you can easily change something. 

As an example, assume that somebody gives you a dataset with 2 variables and ask you to analyse this dataset. With `SPSS` you read this dataset in and then click on various butttons so you get some output. Now, assume further that this person actually has forgotten a variable (this happens more often than you think) and gives you a new dataset with 3 variables. Then you have to do all the clicking again (and hopefully you remembered on which buttons you actually clicked).

When you have a script file you only have to change the code in 1 or 2 places and run it all again. No sweat! So, writing up all the commmands and save it for later, might cost you some time in the **beginning**, but there are **huge** time savers later on! To run a script you simply need to press the button Edit > Run all (`R` editor) or Code > Run Region > Run All (`RStudio`). Nobody does that however, because there are numerous Keyboard Shortcuts (I advise you to learn them, because they make you considerably faster. Actually, most Keyboard Shortcuts work in a wide variety of editors--even in `Word`.)

### Help!

Sometimes you have found a command you would think you could use, but you do not know how, then you need to use `?` operator in front of the command. For instance, you know that the command `c()` could be useful, but how? Then type:

```{r,  eval=FALSE}
?c()
```

And the appropiate documentation will pop up.

### Packages {#subsec:packages}

`R` (as many other software applications nowadays) depends heavily on packages written by other parties, usually users of `R`. There are now many packages out there. You can find the 'official' ones on [the CRAN website](https://cran.r-project.org/web/packages/available_packages_by_name.html), but there are many more. Packages have to be installed (both `R` editor and `RStudio` have a separate package manager) and afterwards loaded. Say, for instance we want to use the awesome `ggplot2` package (a package to make plots look nice, actually to make more elaborate plots but anyway), then to install and load the package we give the commands:

```{r, eval=FALSE}
install.packages("ggplot2")
library("ggplot2")
```

Now, we can use the commands from this package as if they were built-in.

Gradually, we come across some useful packages. Those we will use in a chapter will always be listed at the start of the chapter.

### Using comments

A final word in this section about the use of comments. **Do it!**. Really, it will make your future life much easier if you have documented what you have done. You can insert comments by using the `#` operator (everything after the hashtag is a comment), so, e.g.,

```{r, eval=FALSE}
2+2 # always wanted to know about the outcome, but were afraid to ask
```

## Reading and writing data

To properly do statistics one needs data (duh!). Luckily, there are numerous ways to get data in `R`. 

When you just have a `csv` text file (comma separated file), it is easy, you just type:

```{r, eval=FALSE}
df <- read.csv(file="my_data.csv", header = TRUE)
```

and you read in the `data.csv` in a data frame variable called `df`. Note that the original header variables are preserved. If you would like to store your data you can do the reverse, namely:

```{r, eval=FALSE}
write.csv(df, file = "my_data.csv")
```

Now sometimes you do not have nicely formatted `.csv` or `.txt`files, but nasty `.dta` files from Stata or `.xlsx` files from `Excel`. Here the package Rio comes very handed, being the swiss-army knife of data converters in `R`. Assume you have the `mtcars` dataset in various formats, then you can do 

```{r, eval=FALSE}
library("rio")

x <- import("mtcars.sav")  # SPSS data file
y <- import("mtcars.xlsx") # Excel data file
z <- import("mtcars.dta")  # Stata data file
```

and all dataframes `x`, `y` `z` should be identical.

## Ehmmm, dataframes

I already talked a bit about dataframes, but have not yet explained what they are. Simply, it is your data. Lets construct a simple dataframe:

```{r}
Names  <- c("Thomas", "Erik", "Mark", "Eveline")
Grades <- c(5, 8, 6.5, 7)
Female <- c(FALSE,FALSE,FALSE,TRUE)

df_grades <- data.frame(Names, Grades, Female) 
```

First,note a couple of things. A dataframe can consists of various data_types, in this case strings (the names), numbers (the grades) and so-called Booleans (someone is female or not). Secondly, we have names the variables. So, we now have a dataframe called df_grades. Great, now what? Well, we can do a couple of things. By using the command `head()` we can show the first 6 rows of this dataframe.

```{r}
head(df_grades)
```

Because we only have four observations this actually gives our whole dataframe (you can also just type `df_grades` to get this). Using square brackets `[]` allows you get specific information from this dataset, where the first index denotes the row and the second index denotes the column. Look at the following examples:

```{r}
df_grades[1,2]
df_grades[1,]
df_grades[3]
df_grades["Names"]
```

Most of the statistical stuff we will do invokes the use of dataframe and specific variables from that dataframe. 

## Regression modeling

Before we start laying-out how to do regression modeling in `R`, we first need data. And for this purpose we will simulate our data by the following commands:

```{r}
x <- runif(100, min = 0, max = 1) # create 100 uniformly distributed numbers in interval (0,1)
y <- 2 + 6*x + rnorm(100, mean = 0, sd = 1) # rnorm stands for the normal distribution
df <- data.frame(y,x) # Strictly not necessary but for the sake of the exposition
```

So in fact we have now created the following model:

$$
y_i = 2 + 6 x_i + \epsilon_i,
$$
where $\epsilon_i$ is standard normally distributed (as I will explain later in Section \@ref(sec:theory) this is for convenience but **not** absolutely needed to do linear regression).

checking this with

```{r}
head(df)
```

indeed shows the first 6 combination of $x$ and $y$. If we now perform a linear regression, then we expect that the estimated intercept should be very close to 2 and the estimated slope parameter should be very close to 6. In `R` we have the command `lm()` (from linear model) to do this as follows:

```{r}
linear_model <- lm(y~x, data = df)
```

We now have performed a regression of $y$ onto $x$, using the data `df` and save the result in a variable called linear_model. First note the first part of the `lm(,)` expression. `y~x` means that `y` is the left-hand side variable and `x` the righ-hand side variable (if we have more variables, the formula becomes something as`y~x+u+v+w+z`). The second part of the `lm(,)` expression denotes the specific dataframe to be used. Namely, in `R` you can have multiple dataframes so you have to specify which one is to be used. 

Right, but now what? Well, we have now a variable called linear_model, just typing in `linear_model` only gets you the real basic results:

```{r}
linear_model
```

But, you want more, right? Standard errors, t-statistics, R-squares, the whole lot. For this, you need the `summary()` command, which gives you the following outcome 

```{r}
summary(linear_model)
```

Interestingly, our parameters are close to 2 and 6, but not that close (which happens with a limited amount of observations, notice as well the relatively large standard errors). So, the command `summary()` gives all the needed statistical output, but what about regression diagnostics. For this you can ask for a plot of the variable linear_model

```{r,fig.path='./'}
plot(linear_model)
```

And that is basically it, at least for the basics. There is much more to say about the `lm()` command, but that is for later.  

## Making plots 

Where `R` truly shines is in making plots, diagrams, histograms, etcetera. The first thing with data you want to do is to make a scatterplot. With our $x$ and $y$ data this can be easily done by: 

```{r,fig.path='./'}
plot(x,y)
```

If you would like to create a histogram, just use `hist()` as 

```{r,fig.path='./'}
hist(y)
```

However, to go one step further you also make a plot of a dataframe. for our df dataframe this is not very insightful, but let's add another variable $z$ uncorrelated with $x$ and $y$ and then plot the dataframe (the `$` indicate that `z` is a variable in dataframe `df`).

```{r,fig.path='./'}
df$z <- runif(100, 0,1)
plot(df)
```

And luckily, this plot confirms what we expect. $x$ are correlated $y$ by construction and $z$ is not correlated with either $x$ and $y$. These are all the so-called baseline plots. They are great (and already highly customizable), but lately there has been a new kid on the block called ggplot2. It goes to far to explain the details of ggplot2 (gg here stands for the grammar of graphics), but suffice to say that ggplot2 works with building blocks, so that every piece of the figure that you want (or can think of) can be constructed. Just as an example, let's redo our scatterplot but now using ggplot2 and say we want to add some density lines from our observations (just because we can). This can be done in the following way

```{r,fig.path='./'}
ggplot(df, aes(x,y))+geom_point() + geom_density2d()
```

## Recap

For the absolute beginner `R` is huge and daunting. You need to learn by taking small steps and by practicing (a lot). Do not aim immediately at big and complex projects but start small and at the basics. You will then learn that you quickly make progress and at a certain even become efficienter than when using `Excel` or `SPSS`. In later chapters I dive in to some more detailed topics, and hopefully the material provides you with a background solid enough to understand and work with those topics.  